---
title: "Density estimation. Clustering. (Assignment)"
author: "Pedro Delicado"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(cluster)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, comment = FALSE, error = FALSE, eval=FALSE)
#Put eval=FLSE for having just the question steatments
#Put eval=TRUE for obtaining the output
```

## Reading the data: Boston Housing
We'll use the `MASS::Boston` dataset, that contains median house values from Boston neighbourhoods.
In particular we are interested in the joint distributions of centered and scaled variables `lstat` and `rm`:

```{r,fig.height=4.5,fig.width=4.5,fig.align='center',echo=TRUE,eval=TRUE}
data("Boston", package = "MASS")
X <- scale(Boston[,c(13,6)])
plot(X,as=1)
```

## Questions

1. We want to estimate the joint bivariate density using a kernel estimator with the same bandwith in both dimensions: $h=(a,a)$. 
For instance, the following code performs this estimation for $a=0.5$:

```{r,fig.height=4.5,fig.width=4.5,fig.align='center',echo=TRUE,eval=TRUE}
library(sm)
plot(X,as=1,col=8)
sm.density(X,h=.5*c(1,1),display="slice",props=c(25,50,75,95),col=2,add=TRUE)
```

Use the *maximum log-likelihood cross-validation method* for choosing the value of $a$, whem $a$ takes values in the vector `seq(0.1,1,by=0.1)`. Then repeat the previous density estimation using the chosen value of $a$.

*Indication:* 
Maximize in $a$ the **logarithm** of the likelihood cross-validation (instead of maximizing just the likelihood cross-validation). 
The following code evaluates the logarithm of the density estimator at point $(0,0)$:
```{r,echo=TRUE,eval=FALSE}
new.point <- matrix(c(0,0),ncol=2)
f.hat <- sm.density(X,h=.5*c(1,1),display="none",eval.grid=FALSE,
                    eval.points=new.point)
log(f.hat$estimate)
```


```{r}
va <- seq(0.1,1,by=0.1)
na <- length(va)
logLCVa <- numeric(na)

n <- dim(X)[1]

for (j in 1:na){
  a <- va[j]
  for (i in 1:n){
    new.point <- matrix(X[i,],ncol=2)
    f.hat.i <- sm.density(X[-i,],h=a*c(1,1),display="none",eval.grid=FALSE,
                    eval.points=new.point)$estimate
    logLCVa[j] <- logLCVa[j] + log(f.hat.i)
  }
}

plot(va,logLCVa,type="b")
```

```{r,fig.asp=1}
a.opt <- va[which.max(logLCVa)]

plot(X,as=1,col=8,main=paste("Optimal a: ",a.opt))
sm.density(X,h=a.opt*c(1,1),display="slice",props=c(25,50,75,95),col=2,add=TRUE)
```


2. Do a hierarchical clustering of these data using the `ward.D` method, plot the resulting dendogram and cut it into $k=3$ clusters. 
Plot the scatterplot of the data, using a different color for points in different clusters.

```{r clustering}
k<-3
d.X <- dist(X)
hcl.0 <- hclust(d.X,method = "ward.D")
plot(hcl.0)
hcl.0.k <- cutree(hcl.0,k)
```

```{r,fig.asp=1}
plot(X,col=hcl.0.k,as=1)
```

3. For each one of the $k$ clusters obtained above, do the following tasks *(A unique plot should be done, at which the $k$ densities are represented simultaneously)*:
- Consider the bivariate data set of the points in this cluster.
- Estimate non-parametrically the joint density of `lstat` and `rm`, conditional to this cluster *(Use the optimal bandwith found in the first point)*.
- Represent the estimated bivariate density using the level curve that covers the 75% of the points in this cluster.

 

```{r densities,fig.asp=1,warning=FALSE, message=FALSE}
library(sm)
plot(X,col=hcl.0.k,as=1)
for (j in 1:k){
  cl.j <- (hcl.0.k==j)
  sm.density(X[cl.j,],h=a.opt*c(1,1), 
             display="slice",props=c(75),
             col=j, cex=4, add=TRUE)
}
```

4. Repeat now points 3 and 4, but choose the number of clusters $k$ according to one (or several) of the automatic criteria we have seen in class.
*Optional: If you want, you can choose the optimal bandwidth for each cluster separately (this will improve the final density estimations).* 


```{r clusternum, warning=FALSE, message=FALSE}
library(fpc)
n.cl <- 2:10
l.n.cl<-length(n.cl)

avg.sil <- numeric(l.n.cl)
w <- numeric(l.n.cl)
wb.rat <- numeric(l.n.cl)
sep.ind <- numeric(l.n.cl)
CalHar <- numeric(l.n.cl)

d.Eucl <- dist(Boston[-ncol(Boston)])

logL <- c()

for (i in (1:l.n.cl)){
  k_bar<-n.cl[i]
  hcl.Eucl <- hclust(d.Eucl, method="ward.D")
  cut.hcl.Eucl <- cutree(hcl.Eucl,k_bar)
  cl.stats <- cluster.stats(d.Eucl,cut.hcl.Eucl)
  avg.sil[i]   <- cl.stats$avg.silwidth
  w[i]    <- cl.stats$average.within
  wb.rat[i]    <- cl.stats$wb.ratio
  sep.ind[i]   <- cl.stats$sindex
  CalHar[i]   <- cl.stats$ch
  for (j in 1:k_bar){
    cl.j <- (hcl.0.k==j)
    f.hat.i <- sm.density(X[cl.j,],h=a.opt*c(1,1), 
               display="slice",props=c(75),
               col=j, cex=4)$estimate
    append(logL, log(f.hat.i))
  }
  
}

```




