---
title: 'Assignment: t-SNE'
author: "Kevin Michael Frick, Sonia Petrini, Patrick Lutz"
output: pdf_document
---

# t-SNE for ZERO digits

```{r setup, include=FALSE}
library(Rtsne)
library(dplyr)
library(ggplot2)
library(reshape2)
library(ggpubr)
set.seed(50321)
zip.train <- read.table("zip.train")
zeros <- filter(zip.train, V1==0)
zeros <- as.matrix(zeros[,-1])

# plotting 1 digit
plot.zip <- function(x,use.first=FALSE,...){ x<-as.numeric(x)
if (use.first){
x.mat <- matrix(x,16,16) }else{
x.mat <- matrix(x[-1],16,16) }
image(1:16,1:16,x.mat[,16:1], col=gray(seq(1,0,l=12)),...)
invisible(
if (!use.first){
title(x[1]) }else{
} )
  #col=gray(seq(1,0,l=2)))
}

```

## a

Look for a 2-dimensional ($q = 2$) configuration of the data using parameters `perplexity = 40` and `theta = 0` in `Rtsne` function. Do the scatterplot of the obtained 2-dimensional configuration.

```{r}
n <- dim(zeros)[1]

q<-2 # 2-dim config


Rtsne.zeros.res <- Rtsne(zeros, perplexity=40, theta=0)
conf.Rtsne.zeros.res <- Rtsne.zeros.res$Y
rownames(conf.Rtsne.zeros.res) <- seq(1,nrow(conf.Rtsne.zeros.res))
rownames(zeros) <- seq(1,nrow(zeros))

plot(conf.Rtsne.zeros.res, as=1,
   main="t-SNE, perplexity = 40, theta = 0",
   xlab = "tSNE1", ylab = "tSNE2")
```



## b

In the previous scatterplot, select a few points (9 points, for instance) located in such a way that they cover the variability of all the points in the scatterplot. Then use the function plot.zip to plot the ZERO digits corresponding to the selected points. The images you are plotting should allows you to give an interpretation of the 2 coordinates obtained by t-SNE (observe how the shape of ZEROs changes when moving along each directions of the scatterplot).

### Representative digits

In order to understand the meaning of the two retained dimensions, we select 9 representative examples by considering all 9 combinations of low, around 0, and high values of $tSNE_1$ and $tSNE_2$.
To do so, we select arrays containing the points that lie closest to the axes and then select those that lie at the extremes of these arrays. We do the same with the points which lie near the lines $tSNE_1 = tSNE_2$ and $tSNE_1 = -tSNE_2$. We plot the selected points in purple over the previous scatter plot.

```{r}
ReprPoints <- function(config,refD2,refD1=NULL,setD1=F) {
  config <- data.frame(config) %>% mutate(id = seq(1:nrow(config)))
  colnames(config) <- c("D1","D2","id")
  reflinesD2 <- c("high.D2","null.D2","low.D2")
  pts <- c()
  for (line in reflinesD2) {
    D2condition <- switch(line, "high.D2" = (abs(config$D2 - refD2[1]) < 2),
                                "null.D2" = (abs(config$D2 + 0) < 1),
                                "low.D2"  = (abs(config$D2 + refD2[2]) < 2))
    if (setD1==T) {
      lowD1  <- config[D2condition,] %>% filter(abs(D1 + refD1[2]) < 2) %>%
        slice(1)
      zeroD1 <- config[D2condition,] %>% filter(abs(D1 + 0) < 1) %>% slice(1)
      highD1 <- config[D2condition,] %>% filter(abs(D1 - refD1[1]) < 2) %>%
        slice(1)
      ids <- c(lowD1$id,zeroD1$id,highD1$id)
      pts <- append(pts,ids)
    } else {
      res <- config[D2condition,]
      res <- res %>% arrange(D1) %>% slice(c(1,round(nrow(res)/2),nrow(res)))
      pts <- append(pts,res$id)
    }
  }
  pts
}

PlotWithRepr <- function(config,pts,main="Configuration",xlab="tSNE1",ylab="tSNE2") {
  plot(config, as=1,
     main=main,
     xlab = xlab, ylab = ylab,
     col= ifelse(rownames(config) %in% pts,"magenta","black"),
     pch= ifelse(rownames(config) %in% pts,19,1),
     cex= ifelse(rownames(config) %in% pts,2,1))
}

pts <- ReprPoints(conf.Rtsne.zeros.res,c(20, 20))
PlotWithRepr(conf.Rtsne.zeros.res, pts,
             main = "t-SNE, perplexity = 40, theta = 0")

```


```{r}
PlotReprZeros <- function(pts, x1 = "tSNE1", x2 = "tSNE2"){
  mains <- c(paste("High ",x2, ", Low ",x1, sep=""),
             paste("High ",x2, ", Around zero ", x1, sep=""),
             paste("High ",x2, ", High ", x1, sep=""),
             paste("Around zero ",x2, ", Low ", x1, sep=""),
             paste("Around zero ",x2, ", Around zero ", x1, sep=""),
             paste("Around zero ",x2, ", High ", x1, sep=""),
             paste("Low ",x2, ", Low ", x1, sep=""),
             paste("Low ",x2, ", Around zero ", x1, sep=""),
             paste("Low ",x2, ", High " , x1, sep=""))

  p <- par(mfrow=c(3,3))
  for (i in 1:length(pts)) {
    point <- pts[i]
    plot.zip(zeros[point,],TRUE,main=mains[i])
  }
  par(p)
}

PlotReprZeros(pts)
```


By plotting the representative datapoints, we can get an idea of the meaning of the two dimensions $tSNE_1$ and $tSNE_2$.
The dimension $tSNE_1$ seems to encode a notion of radius of the zero, with higher values corresponding to wider zeros. When $tSNE_1$ is around zero, $tSNE_2$ seems to be positively correlated with stroke width, but such a correlation seems to be in the opposite direction when $tSNE_1$ has very positive or negative values. In general, $tSNE_2$ seems to be related with an interplay of characteristics of the zeroes.



## c (OPTIONAL)

Relate the results from t-SNE with those obtained by the first 3 principal components. In particular, could you represent in any way the results obtained by t-SNE in the 3-dimensional scatter plot of (PC1,PC2,PC3)?

We perform PCA on our dataset and plot the configuration obtained with the first two dimensions. Again, we pick some representative points to understand the underlying concepts.

```{r}
zeros.pca <- princomp(as.matrix(zeros))
pc1 <- zeros.pca$scores[,1]
pc2 <- zeros.pca$scores[,2]
pc3 <- zeros.pca$scores[,3]

pts12 <- ReprPoints(cbind(pc1, pc2),c(7,7))
pts23 <- ReprPoints(cbind(pc2, pc3),c(7,5),c(7,7),setD1 = T)
pts31 <- ReprPoints(cbind(pc3, pc1),c(7,7),c(7,5),setD1 = T)
par(mfrow=c(1,3))
PlotWithRepr(cbind(pc1, pc2),pts12,main="PCA configuration, PC1 and PC2",xlab="PC1",ylab="PC2")
PlotWithRepr(cbind(pc2, pc3),pts23,main="PCA configuration, PC2 and PC3",xlab="PC2",ylab="PC3")
PlotWithRepr(cbind(pc3, pc1),pts31,main="PCA configuration, PC3 and PC1",xlab="PC3",ylab="PC1")
```

### Representative digits

```{r}
PlotReprZeros(pts12, "PC1", "PC2")
PlotReprZeros(pts23, "PC2", "PC3")
PlotReprZeros(pts31, "PC3", "PC1")
```
The first principal component seems to be related to the flatness of the zero, but the direction of the correlation is the opposite with respect to what is expressed by t-SNE: higher $PC1$ values implying a thinner digit. The second principal component is  an indicator of the stroke width, and it moves in a positive direction. In both cases, a low value of $PC3$ seems to reduce the influence of $PC1$ and $PC2$ and make for "more equal" zeros, as if $PC3$ encoded a notion of "variance" of the points.

### Dimensions Correlation

The only immediately apparent relationship between t-SNE and PCA is the negative correlation between the first t-SNE component and $PC1$. Therefore, we compute the correlations between them, and display their scatter plots.

```{r}
tSNE1 <- conf.Rtsne.zeros.res[,1]
tSNE2 <- conf.Rtsne.zeros.res[,2]
df <- data.frame(pc1,pc2,pc3,tSNE1,tSNE2)

pl1 <- df %>% ggplot(aes(pc1,tSNE1)) + geom_point()
pl2 <- df %>% ggplot(aes(pc3,tSNE2)) + geom_point()
pl3 <- df %>% cor() %>% round(2) %>% melt() %>%
ggplot(aes(x=Var1, y=Var2, fill=value, label = value)) +
  geom_tile() + geom_text() + scale_fill_gradient2(low = "blue", high = "red", mid = "white",
   midpoint = 0, limit = c(-1,1), space = "Lab",
   name="Corr")

ggarrange(pl3,ggarrange(pl1,pl2,nrow=2))
```

As expected from visualizing the representative digits, the two first components $PC_1$ and $tSNE_1$ are strongly negatively related. Moreover, there seems to be a positive correlation, albeit weaker, between $PC_3$ and $tSNE_2$, as well as between $PC_2$ and $tSNE_2$ and, to a much lower extent, between $tSNE_2$ and $PC_1$. This is consistent with the observation that $tSNE_2$ seems to be related to an interplay of characteristics.

### 2-D configuration comparison

We plot the configurations obtained with PCA ($PC_1$ and $PC_3$, which display the strongest correlation, respectively with $tSNE_1$ and $t_SNE_2$) and with t-SNE. t-SNE in this case does a better job at recovering a 2-dimensional representation of the data that explains more variability, albeit the second dimension is harder to interpret compared to PCA. The first t-SNE coordinate clearly explains more variability than the first principal component. This is consistent with what we observed above by visualizing some representative digits: the differences within t-SNE embedded digits across the first dimension where much more evident than those in the PCA ones.

```{r}

pl_pc <- ggplot(df, aes(pc1,pc3)) + geom_point() + xlim(c(-30,30)) + labs(title="PCA configuration")
pl_Rtsne <- ggplot(df, aes(tSNE1,tSNE2)) + geom_point() + xlim(c(-30,30)) + labs(title="t-SNE configuration")
ggarrange(pl_pc,pl_Rtsne,nrow=2)
```


### 2D and 3D configurations

We plot the 3D configuration obtained by PCA.

```{r}
library(scatterplot3d)
scatterplot3d(x = pc1, y = pc2, z = pc3)
```

We choose to fit three linear models that regress each first three principal components on a spline that fits the two components obtained with t-SNE as its covariates. This results into the following expression for each component: $\hat{pc_j} = b_0 + b_1 S(tSNE_1, tSNE_2), \;j \in \{1, 2, 3\}$.
We plot the scatterplot of these linear models over that of the three principal components.

```{r}
library(mgcv)
df.PCA.tSNE <- as.data.frame(cbind(pc1, pc2, pc3, tSNE1, tSNE2))
smooth.PC1 <- gam(pc1~s(tSNE1,tSNE2), data=df.PCA.tSNE)$fitted.values
smooth.PC2 <- gam(pc2~s(tSNE1,tSNE2), data=df.PCA.tSNE)$fitted.values
smooth.PC3 <- gam(pc3~s(tSNE1,tSNE2), data=df.PCA.tSNE)$fitted.values
points1 <- scatterplot3d(pc1, pc2, pc3)
points1$points3d(smooth.PC1,smooth.PC2,smooth.PC3,col=6)
```
# Selecting the tuning parameters for ZERO digits

## a

Use the local continuity meta criteria to select the tuning parameter `perplexity` in t-SNE (use `theta = 0`) for ZERO digits. Then describe graphically the low dimensional configuration corresponding to the optimal parameter.


```{r}
LCMC <- function(D1,D2,Kp){
  D1 <- as.matrix(D1)
  D2 <- as.matrix(D2)
  n <- dim(D1)[1]
  N.Kp.i <- numeric(n)
  for (i in 1:n){
    N1.i <- sort.int(D1[i,],index.return = TRUE)$ix[1:Kp]
    N2.i <- sort.int(D2[i,],index.return = TRUE)$ix[1:Kp]
    N.Kp.i[i] <- length(intersect(N1.i, N2.i))
  }
  N.Kp<-mean(N.Kp.i)
  M.Kp.adj <- N.Kp/Kp - Kp/(n-1)

  return(list(N.Kp.i=N.Kp.i, M.Kp.adj=M.Kp.adj))
}

q <- 2

Kp <- 10

v.perp <- c(10, 15, 20, 25, 40, 60)

LC <- numeric(length(v.perp))
Rtsne.perp <- vector("list",length(v.perp))
D1 <- dist(zeros)
for (i in 1:length(v.perp)){
  print(v.perp[i])
  Rtsne.perp[[i]] <- Rtsne(zeros, ndim=q,perplexity=v.perp[i], theta=0)

  D2.perp <- dist(Rtsne.perp[[i]]$Y)
  LC[i] <- LCMC(D1,D2.perp,Kp)$M.Kp.adj
  #print(c(i,j,LC[i,j]))
}


i.max <- which.max(LC)
perp.max <- v.perp[i.max]
Rtsne.max <- Rtsne.perp[[i.max]]$Y

plot(v.perp, LC, type="b", main=paste0("The perplexity maximizing the LCMC is ",perp.max))
abline(v=perp.max,lty="dashed",col="red")

```


## Best configuration

We now plot the resulting Tuned t-SNE.

```{r}
par(mfrow=c(1,2))
plot(Rtsne.max, main=paste("Tuned t-SNE, perplexity =", perp.max),col="#cc3366", xlab = "tSNE1", ylab = "tSNE2")
plot(conf.Rtsne.zeros.res, as=1, main="t-SNE, perplexity = 40",
     col="darkgreen", xlab = "tSNE1", ylab = "tSNE2")
```

 By tuning the parameter `perplexity` through the local continuity meta criteria, we identify 20 as the optimal number value. We plot the resulting configuration, together with the initial one, obtained with `perplexity = 40` . As we can see, with the tuned parameter both dimensions allow one to explain a greater portion of the data variability, compared to the first estimated configuration with an arbitrary `perplexity`: in fact, the observations now spread wider both on the x and the y axis. However, the overall shape of the data remains the same.


### Real vs Estimated distances

Then, we display a scatter plot of the real distances against the best estimated ones.

```{r}
Rtsne.dist.best  <- sort(dist(Rtsne.max))
Rtsne.dist.perp40 <- sort(dist(conf.Rtsne.zeros.res))
real.dist <- sort(dist(zeros))
sample <- sample(length(real.dist),10000)

plot(Rtsne.dist.best[sample],real.dist[sample],
     main="t-SNE: real vs est. dist. with perplexity = 40",xlab="t-SNE distance",
     ylab="real distance")
plot(Rtsne.dist.perp40[sample],real.dist[sample],
     main=paste0("t-SNE: real vs est. dist. with best perplexity = ", perp.max),
     xlab="t-SNE distance", ylab="real distance")

```

As we can see, t-SNE managed to recover the original distances so as to retain their natural increasing order. As a matter of fact, real distances show a higher increase with respect to the estimated ones, especially for low values. We see that there is not a significant difference in overall shape of the real vs estimated curve, using the optimal `perplexity` value, but we notice that the scale of the distances is slightly lower when using the optimal value.
